Automated residual plot assessment with the \textsf{R} package \textsf{autovi} and the \textsf{Shiny} app \textsf{autovi.web}

### 2025-09-04 autovi_paper.qmd
Fix math notaion and some packages

---
title: "Automated Residual Plot Assessment with the R Package autovi and the Shiny App autovi.web"
format:
  quarto-anzjs-pdf:
    keep-tex: true  
    shorttitle: "Automated Residual Plot Assessment"
    double-space: true
    line-numbers: true
    # You can specify extra preamble in the file below.
    include-in-header:
      - header.tex
date: today
date-format: "YYYY" # used to define volumeyear
author:
  - name: Weihao Li
    email: patrick.li@anu.edu.au
    corresponding: true
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
      - name: The Australian National University
        department: Biological Data Science Institute
        address: 46 Sullivan's Creek Road, ACT 2600, Australia
    orcid: 0000-0003-4959-106X
  - name: Dianne Cook
    email: dicook@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0002-3813-7155
  - name: Emi Tanaka
    email: emi.tanaka@anu.edu.au
    affiliations: 
      - name: The Australian National University
        department: Biological Data Science Institute
        address: 46 Sullivan's Creek Road, ACT 2600, Australia
    orcid: 0000-0002-1455-259X
  - name: Susan VanderPlas
    email: emi.tanaka@anu.edu.au
    affiliations:
      - name: University of Nebraska
        department: Department of Statistics
        address: Hardin Hall, 3310 Holdrege St Suite 340, Lincoln, NE 68583, United States
    orcid: 0000-0002-3803-0972
  - name: Klaus Ackermann
    email: Klaus.Ackermann@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0001-7693-8538
# Note that the Journal requires that a paper must begin with a
# "Summary" not an "Abstract".  This is automatically taken care
# of by the anzsauth document style.  So even though the following
# says "abstract" the heading "Summary" will appear in
# the processed version.
abstract: |
  Visual assessment of residual plots is a common approach for diagnosing linear models, but it relies on manual evaluation, which does not scale well and can lead to inconsistent decisions across analysts. The lineup protocol, which embeds the observed plot among null plots, can reduce subjectivity but requires even more human effort. In today’s data-driven world, such tasks are well-suited for automation. We present a new \textsf{R} package that uses a computer vision model to automate the evaluation of residual plots. An accompanying \textsf{Shiny} app is provided for ease of use. Given a sample of residuals, the model predicts a visual signal strength (VSS) and offers supporting information to help analysts assess model fit. 
# Note that "keywords" should not include words and phrases
# that form part of the title of the paper.
# Also keywords (with the exception of proper names and
# certain abbreviations that conventionally appear all in
# capital letters) should *not* be capitalised.
keywords: [initial data analysis, statistical graphics, data visualization, visual inference, computer vision, machine learning, hypothesis testing, regression analysis, model diagnostics]
bibliography: bibliography.bib  
#nocite: |
---

@@

To make the statistical testing procedure and trained computer vision model widely accessible, we developed the R package \textsf{autovi} along with a companion web interface, \textsf{autovi.web}, which allows users to automatically assess their residual plots using the trained computer vision model.

@@

In the context of regression model diagnostics, VSS describes the clarity of visual patterns on a diagnostic plot that may indicate model violations. Violations can be categorized as weak, moderate, or strong, but here we treat it as a continuous positive real variable. Importantly, its interpretation depends on how it is linked to a function of the data or the underlying data generating process. Consequently, the calculation of VSS can vary across different model classes or within the same model, depending on the generating function.

@@

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\boldsymbol{e})}{q(\boldsymbol{e})}p(\boldsymbol{e})d\boldsymbol{e},
\end{equation}

here, $p(.)$ and $q(.)$ are the probability density functions of the reference residual distribution $P$ and the true residual distribution $Q$, respectively.

This distance measure depends on knowledge of the true residual distribution, which is unknown in practice. To compute $D_{KL}$ for the training samples, Equation \ref{eq:kl-1} takes different forms depending on the specific model violations. For instance, where necessary higher-order predictors, $\boldsymbol{Z}$, and their corresponding parameter, $\boldsymbol{\beta}_Z$, are omitted from the fitted linear model, the distance measure can be expanded as follows:

@@

# R package: autovi {#sec-autovi}

The main purpose of `autovi` is to provide rejection decisions and $p$-values for testing the null hypothesis ($H_0$) that the regression model is correctly specified. The package provides automated interpretation of residual plots using computer vision. The name `autovi` stands for **auto**mated **v**isual **i**nference. This functionality can be accessed through the R package `autovi`, or through a web interface, `autovi.web`, which enables users to perform analyses without installing R, Python, or their associated dependencies locally.

@@

The `autovi` infrastructure effectively integrates multiple programming languages and libraries into a comprehensive analytical tool. It relies on five core libraries from Python and R, each playing a critical role in the analysis pipeline. In Python, `pillow` [@clark2015pillow] handles image processing tasks such as reading and resizing PNG files of residual plots, then converting them into input tensors for further analysis. `TensorFlow` [@abadi2016tensorflow], a key component of modern machine learning, is used to predict the VSS of these plots using a pre-trained convolutional neural network.

In the R environment, `autovi` utilizes several libraries. `ggplot2` [@ggplot2] generates the initial residual plots, saved as PNG files for visual input. `cassowaryr` [@mason2022cassowaryr] computes scagnostics (scatter plot diagnostics), providing numerical features that capture statistical properties of the plots. These scagnostics complement the visual analysis by offering quantitative metrics as secondary input to the computer vision model. `reticulate` [@reticulate]  enables seamless communication between R and Python.

@@

The `autovi` package is available on CRAN. It is actively developed and maintained, with the latest updates accessible on GitHub. This paper uses `autovi` version 0.4.2.
The package includes internal functions to check the current Python environment used by the `reticulate` package. If the necessary Python packages are not installed in the Python interpreter, an error will be raised. If you want to select a specific Python environment, you can do so by calling the `reticulate::use_python()` function before using the `autovi` package.

We recommend using the Shiny app `autovi.web` if users encounter installation problems.

@@

```{r out.width = "100%"}
#| label: fig-autovi-diag
#| fig-cap: "Diagram illustrating the infrastructure of the R package \\texttt{autovi}. The modules in green are primary inputs provided by users. Modules in blue are overridable methods that can be modified to accommodate users' specific needs. The module in yellow is a pre-defined non-overridable method. The modules in red are primary outputs of the package."
magick::image_read_pdf("figures/autovi.pdf")
```

@@

The `autovi` R package can be extended to accommodate other classes of models beyond linear regression, such as generalized linear models (`glm`). This is achieved by substituting the relevant overridable modules, and if needed, supplying a different `Keras` model.

@@

The `autovi.web` shiny application extends the functionality of `autovi` by offering a user-friendly web interface for automated residual plot assessment. This eliminates the common challenges associated with software installation, so users can avoid managing Python environments or handling version requirements for R libraries. The platform is cross-platform and accessible on various devices and operating systems, making it suitable even for users without R programming experience. Additionally, updates are managed centrally, ensuring that users always have access to the latest features. This section discusses the implementation based on `autovi.web` version 0.1.0.

@@

The interface `autovi.web` is built using the `shiny` [@shiny] and `shinydashboard` [@shinydashboard] R packages. Hosted on the [shinyapps.io](https://www.shinyapps.io) domain, the application is accessible through any modern web browser. The R packages `htmltools` [@htmltools] and `shinycssloaders` [@shinycssloaders] are used to render markdown documentation in shiny application, and for loading animations for shiny widgets, respectively.

@@

We also explored the possibility of implementing the web interface using frameworks built on other languages, such as Python. However, server hosting domains that natively support Python servers typically do not have the latest version of R installed. Additionally, calling R from Python is typically done using the `rpy2` Python library [@rpy2], but this approach can be awkward when dealing with language syntax related to non-standard evaluation. Another option we considered was renting a server where we could have full control, such as those provided by cloud platforms like Google Cloud Platform (GCP) or Amazon Web Services (AWS). However, deploying and maintaining the server securely requires some expertise. Ultimately, the most practical solution was to use the `shiny` and `shinydashboard` frameworks, which are well-established in the R community and offer a solid foundation for web application development.

The server-side configuration of `autovi.web` is carefully designed to support its functionality. Most required Python libraries, including `pillow` and `numpy`, are pre-installed on the server. These libraries are integrated into the Shiny application using the `reticulate` package, which provides an interface between R and Python.

Due to shinyapps.io's resource policy, inactive servers enter sleep mode, clearing the local Python environment. When reactivated for a new session, libraries must be reinstalled. While this ensures a clean environment for each session, it may lead to slightly longer loading times for the first user after a period of inactivity.

@@

While `autovi` requires downloading the pre-trained computer vision models from GitHub, these models in ".keras" file format are incompatible with `TensorFlow.js`. Therefore, we extract and store the model weights in JSON files and include them as extra resources in the Shiny application. When the application initializes, `TensorFlow.js` rebuilds the computer vision model using these pre-stored weights.

To allow communication between `TensorFlow.js` and other components of the Shiny application, the `shinyjs` R package [@shinyjs] is used. This package allows calling custom JavaScript code within the Shiny framework. The specialized JavaScript code for initializing `TensorFlow.js` and calling `TensorFlow.js` for VSS prediction is deployed alongside the Shiny application as additional resources.

@@

The workflow of `autovi.web` is designed to be straightforward, with numbered steps displayed in each panel. There are two example datasets provided by the web application. The single residual plot example uses the `dino` dataset from the R package `datasauRus` [@datasaurus]. The lineup example uses residuals from a simulated regression model that has a non-linearity issue. We walk through the lineup example to further demonstrate the workflow of the web application.

@@

This paper presents new regression diagnostics software, the R package `autovi` and its accompanying web interface, `autovi.web`. It addresses a critical gap in the current landscape of statistical software. While regression tools are widely available, effective and efficient diagnostic methods have lagged behind, particularly in the field of residual plot interpretation.

The `autovi` R package, introduced in this paper, automates the assessment of residual plots by incorporating a computer vision model, reducing reliance on time-consuming and potentially inconsistent human interpretation. This automation improves the efficiency of the diagnostic process and promotes consistency in model evaluation across different users and studies.

The development of the accompanying Shiny app, `autovi.web`, expands access to these advanced diagnostic tools, by providing a user-friendly interface. It makes automated residual plot assessment accessible to a broader audience, including those who may not have extensive programming experience. This web-based solution effectively addresses the potential barriers to adoption, such as complex dependencies and installation requirements, that are often associated with advanced statistical software.

@@

The current version of `autovi` can be installed from CRAN, and source code for both packages are available at [`github.com/TengMCing/autovi`](https://github.com/TengMCing/autovi) and [`github.com/TengMCing/autovi_web`](https://github.com/TengMCing/autovi_web) respectively. The web interface is available from [`autoviweb.netlify.app`](https://autoviweb.netlify.app/).

@@

These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor], `glue` [@glue],  `here` [@here], `magick` [@magick], `yardstick` [@yardstick] and `reticulate` [@reticulate].

>>>>

---
title: Automated residual plot assessment with the \textsf{R} package \textsf{autovi} and the \textsf{Shiny} app \textsf{autovi.web}
format:
  quarto-anzjs-pdf:
    keep-tex: true  
    shorttitle: "Automated Residual Plot Assessment"
    double-space: true
    line-numbers: true
    # You can specify extra preamble in the file below.
    include-in-header:
      - header.tex
date: today
date-format: "YYYY" # used to define volumeyear
author:
  - name: Weihao Li
    email: patrick.li@anu.edu.au
    corresponding: true
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
      - name: The Australian National University
        department: Biological Data Science Institute
        address: 46 Sullivan's Creek Road, ACT 2600, Australia
    orcid: 0000-0003-4959-106X
  - name: Dianne Cook
    email: dicook@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0002-3813-7155
  - name: Emi Tanaka
    email: emi.tanaka@anu.edu.au
    affiliations: 
      - name: The Australian National University
        department: Biological Data Science Institute
        address: 46 Sullivan's Creek Road, ACT 2600, Australia
    orcid: 0000-0002-1455-259X
  - name: Susan VanderPlas
    email: emi.tanaka@anu.edu.au
    affiliations:
      - name: University of Nebraska
        department: Department of Statistics
        address: Hardin Hall, 3310 Holdrege St Suite 340, Lincoln, NE 68583, United States
    orcid: 0000-0002-3803-0972
  - name: Klaus Ackermann
    email: Klaus.Ackermann@monash.edu 
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: Wellington Road, VIC 3800, Australia
    orcid: 0000-0001-7693-8538
# Note that the Journal requires that a paper must begin with a
# "Summary" not an "Abstract".  This is automatically taken care
# of by the anzsauth document style.  So even though the following
# says "abstract" the heading "Summary" will appear in
# the processed version.
abstract: |
  Visual assessment of residual plots is a common approach for diagnosing linear models, but it relies on manual evaluation, which does not scale well and can lead to inconsistent decisions across analysts. The lineup protocol, which embeds the observed plot among null plots, can reduce subjectivity but requires even more human effort. In today’s data-driven world, such tasks are well-suited for automation. We present a new \textsf{R} package that uses a computer vision model to automate the evaluation of residual plots. An accompanying \textsf{Shiny} app is provided for ease of use. Given a sample of residuals, the model predicts a visual signal strength (VSS) and offers supporting information to help analysts assess model fit. 
# Note that "keywords" should not include words and phrases
# that form part of the title of the paper.
# Also keywords (with the exception of proper names and
# certain abbreviations that conventionally appear all in
# capital letters) should *not* be capitalised.
keywords: [initial data analysis, statistical graphics, data visualization, visual inference, computer vision, machine learning, hypothesis testing, regression analysis, model diagnostics]
bibliography: bibliography.bib  
#nocite: |
---

@@

To make the statistical testing procedure and trained computer vision model widely accessible, we developed the \textsf{R} package \textsf{autovi} along with a companion web interface, \textsf{autovi.web}, which allows users to automatically assess their residual plots using the trained computer vision model.

@@

In the context of classical normal linear regression model diagnostics, VSS describes the clarity of visual patterns on a diagnostic plot that may indicate model violations. Violations can be categorized as weak, moderate, or strong, but here we treat it as a continuous positive real variable. Importantly, its interpretation depends on how it is linked to a function of the data or the underlying data generating process. Consequently, the calculation of VSS can vary across different model classes or within the same model, depending on the generating function.

@@

\begin{equation} \label{eq:kl-1}
D_{KL} = \int_{\mathbb{R}^{n}}\log\frac{p(\hat{\boldsymbol{e}})}{q(\hat{\boldsymbol{e}})}p(\hat{\boldsymbol{e}})d\hat{\boldsymbol{e}},
\end{equation}

here, $\hat{\boldsymbol{e}}$ denotes the residual vector from the regression model, and $p(\cdot)$ and $q(\cdot)$ are the probability density functions of the reference residual distribution $\mathrm{P}$ and the true residual distribution $\mathrm{Q}$, respectively.

This distance measure depends on knowledge of the true residual distribution, which is unknown in practice. To compute $D_{KL}$ for the training samples, Equation (\ref{eq:kl-1}) takes different forms depending on the specific model violations. For instance, where necessary higher-order predictors, $\boldsymbol{Z}$, and their corresponding parameter, $\boldsymbol{\beta}_Z$, are omitted from the fitted linear model, the distance measure can be expanded as follows:

@@

# \textsf{R} package: autovi {#sec-autovi}

The main purpose of `autovi` is to provide rejection decisions and $p$-values for testing the null hypothesis ($H_0$) that the regression model is correctly specified. The package provides automated interpretation of residual plots using computer vision. The name `autovi` stands for **auto**mated **v**isual **i**nference. This functionality can be accessed through the \textsf{R} package `autovi`, or through a web interface, `autovi.web`, which enables users to perform analyses without installing \textsf{R}, \textsf{Python}, or their associated dependencies locally.

@@

The `autovi` infrastructure effectively integrates multiple programming languages and libraries into a comprehensive analytical tool. It relies on five core libraries from \textsf{Python} and \textsf{R}, each playing a critical role in the analysis pipeline. In \textsf{Python}, `pillow` [@clark2015pillow] handles image processing tasks such as reading and resizing PNG files of residual plots, then converting them into input tensors for further analysis. `TensorFlow` [@abadi2016tensorflow], a key component of modern machine learning, is used to predict the VSS of these plots using a pre-trained convolutional neural network.

In the \textsf{R} environment, `autovi` utilizes several libraries. `ggplot2` [@ggplot2] generates the initial residual plots, saved as PNG files for visual input. `cassowaryr` [@mason2022cassowaryr] computes scagnostics (scatter plot diagnostics), providing numerical features that capture statistical properties of the plots. These scagnostics complement the visual analysis by offering quantitative metrics as secondary input to the computer vision model. `reticulate` [@reticulate]  enables seamless communication between \textsf{R} and \textsf{Python}.

@@

The `autovi` package is available on CRAN. It is actively developed and maintained, with the latest updates accessible on GitHub. This paper uses `autovi` version 0.4.2.
The package includes internal functions to check the current \textsf{Python} environment used by the `reticulate` package. If the necessary \textsf{Python} packages are not installed in the \textsf{Python} interpreter, an error will be raised. If you want to select a specific \textsf{Python} environment, you can do so by calling the `reticulate::use_python()` function before using the `autovi` package.

We recommend using the \textsf{Shiny} app `autovi.web` if users encounter installation problems.

@@

```{r out.width = "100%"}
#| label: fig-autovi-diag
#| fig-cap: "Diagram illustrating the infrastructure of the \textsf{R} package \\texttt{autovi}. The modules in green are primary inputs provided by users. Modules in blue are overridable methods that can be modified to accommodate users' specific needs. The module in yellow is a pre-defined non-overridable method. The modules in red are primary outputs of the package."
magick::image_read_pdf("figures/autovi.pdf")
```

@@

The `autovi` \textsf{R} package can be extended to accommodate other classes of models beyond linear regression, such as generalized linear models (`glm`). This is achieved by substituting the relevant overridable modules, and if needed, supplying a different `Keras` model.

@@

The `autovi.web` \textsf{Shiny} application extends the functionality of `autovi` by offering a user-friendly web interface for automated residual plot assessment. This eliminates the common challenges associated with software installation, so users can avoid managing \textsf{Python} environments or handling version requirements for \textsf{R} libraries. The platform is cross-platform and accessible on various devices and operating systems, making it suitable even for users without \textsf{R} programming experience. Additionally, updates are managed centrally, ensuring that users always have access to the latest features. This section discusses the implementation based on `autovi.web` version 0.1.0.

@@

The interface `autovi.web` is built using the `shiny` [@shiny] and `shinydashboard` [@shinydashboard] \textsf{R} packages. Hosted on the [shinyapps.io](https://www.shinyapps.io) domain, the application is accessible through any modern web browser. The \textsf{R} packages `htmltools` [@htmltools] and `shinycssloaders` [@shinycssloaders] are used to render markdown documentation in \textsf{Shiny} application, and for loading animations for \textsf{Shiny} widgets, respectively.

@@

We also explored the possibility of implementing the web interface using frameworks built on other languages, such as \textsf{Python}. However, server hosting domains that natively support \textsf{Python} servers typically do not have the latest version of \textsf{R} installed. Additionally, calling \textsf{R} from \textsf{Python} is typically done using the `rpy2` \textsf{Python} library [@rpy2], but this approach can be awkward when dealing with language syntax related to non-standard evaluation. Another option we considered was renting a server where we could have full control, such as those provided by cloud platforms like Google Cloud Platform (GCP) or Amazon Web Services (AWS). However, deploying and maintaining the server securely requires some expertise. Ultimately, the most practical solution was to use the `shiny` and `shinydashboard` frameworks, which are well-established in the \textsf{R} community and offer a solid foundation for web application development.

The server-side configuration of `autovi.web` is carefully designed to support its functionality. Most required \textsf{Python} libraries, including `pillow` and `numpy`, are pre-installed on the server. These libraries are integrated into the \textsf{Shiny} application using the `reticulate` package, which provides an interface between \textsf{R} and \textsf{Python}.

Due to shinyapps.io's resource policy, inactive servers enter sleep mode, clearing the local \textsf{Python} environment. When reactivated for a new session, libraries must be reinstalled. While this ensures a clean environment for each session, it may lead to slightly longer loading times for the first user after a period of inactivity.

@@

While `autovi` requires downloading the pre-trained computer vision models from GitHub, these models in ".keras" file format are incompatible with `TensorFlow.js`. Therefore, we extract and store the model weights in JSON files and include them as extra resources in the \textsf{Shiny} application. When the application initializes, `TensorFlow.js` rebuilds the computer vision model using these pre-stored weights.

To allow communication between `TensorFlow.js` and other components of the \textsf{Shiny} application, the `shinyjs` \textsf{R} package [@shinyjs] is used. This package allows calling custom JavaScript code within the \textsf{Shiny} framework. The specialized JavaScript code for initializing `TensorFlow.js` and calling `TensorFlow.js` for VSS prediction is deployed alongside the \textsf{Shiny} application as additional resources.

@@

The workflow of `autovi.web` is designed to be straightforward, with numbered steps displayed in each panel. There are two example datasets provided by the web application. The single residual plot example uses the `dino` dataset from the \textsf{R} package `datasauRus` [@datasaurus]. The lineup example uses residuals from a simulated regression model that has a non-linearity issue. We walk through the lineup example to further demonstrate the workflow of the web application.

@@

This paper presents new regression diagnostics software, the \textsf{R} package `autovi` and its accompanying web interface, `autovi.web`. It addresses a critical gap in the current landscape of statistical software. While regression tools are widely available, effective and efficient diagnostic methods have lagged behind, particularly in the field of residual plot interpretation.

The `autovi` \textsf{R} package, introduced in this paper, automates the assessment of residual plots by incorporating a computer vision model, reducing reliance on time-consuming and potentially inconsistent human interpretation. This automation improves the efficiency of the diagnostic process and promotes consistency in model evaluation across different users and studies.

The development of the accompanying \textsf{Shiny} app, `autovi.web`, expands access to these advanced diagnostic tools, by providing a user-friendly interface. It makes automated residual plot assessment accessible to a broader audience, including those who may not have extensive programming experience. This web-based solution effectively addresses the potential barriers to adoption, such as complex dependencies and installation requirements, that are often associated with advanced statistical software.

@@

The current version of \textsf{autovi} can be installed from CRAN, and source code for both packages are available at [`github.com/TengMCing/autovi`](https://github.com/TengMCing/autovi) and [`github.com/TengMCing/autovi_web`](https://github.com/TengMCing/autovi_web) respectively. The web interface is available from [`autoviweb.netlify.app`](https://autoviweb.netlify.app/).

@@

These \textsf{R} packages were used for the work: \textsf{tidyverse} [@tidyverse], \textsf{lmtest} [@lmtest], \textsf{kableExtra} [@kableextra], \textsf{patchwork} [@patchwork], \textsf{rcartocolor} [@rcartocolor], \textsf{glue} [@glue],  \textsf{here} [@here], \textsf{magick} [@magick], \textsf{yardstick} [@yardstick] and \textsf{reticulate} [@reticulate].
